SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-07 method.tex
finished pass

\subsection{Performance Model}
\label{Method-Model}
For testing and comparing PPF, we use the ChampSim~\cite{Champsim} simulator.
ChampSim is an enhanced version of the framework that was used for the 2nd
Data Prefetch Championship (DPC-2)~\cite{DPC_2}. We model 1-core, 4-core, and
8-core out-of-order machines. The details of the configuration parameters are
summarized in Table~\ref{tab:Sim_params}.

@@

The block size is fixed at 64 bytes. Prefetching is only triggered at an L2
cache demand accesses but could be directed to the L2 or last-level cache. No
L1 data level prefetching is done. The LRU replacement policy is used on all
levels of cache hierarchies. Branch prediction is done using the perceptron
branch predictor~\cite{PerceptronPredictor}. The page size is 4KB. 
ChampSim operates all the prefetchers strictly in the physical address space.

@@

%This is the first time that SPEC CPU 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
%We use the SPEC CPU 2017 benchmark suite~\cite{SPEC2017}. 
We use all the 20 workloads available in the SPEC CPU 2017 suite~\cite{SPEC2017}. Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.\\
%
\noindent \textit{Single-core performance:} For single-core simulations, we use 200
million instructions to warm-up the microarchitectural structures and the next
one billion instructions to do detailed simulations and collect run-time
statistics. We report the IPC speedup over the baseline of no prefetching.
The final numbers reported are the geometric mean of the weighted mean speedup 
achieved per application using the SimPoint methodology.\\
%
\noindent \textit{Multi-core performance:} For multi-application workloads, we generate
100 random mixes and another 100 mixes from the memory intensive subset of
SPEC CPU 2017. For 4-core workload, 200 Million instructions are used for warm-up
and additional 1 Billion instruction simulated for collecting statistics.
Each CPU keeps executing its workload till the last CPU completes one billion
instructions after warm-up. For collecting IPC and other data, only the first
billion instructions are considered as the region of interest.

@@

We repeat the same process for 8-core workloads, correspondingly with 16MB
LLC. The only difference is that 20 million warm-up instructions and 100
million full instructions are executed. This is done so as to keep the
simulation run-time within reasonable limits as a single 8-core mix takes up
to 3 days to simulate one billion instructions.\\
%
\noindent \textit{Validation:} We cross-validated our PPF model using SPEC
CPU 2006~\cite{SPEC2006} and CloudSuite~\cite{CloudSuite} benchmarks. For
single-core SPEC CPU 2006, we developed 94 simpoints spread across all the 29
applications. For multi-core, we followed the same methodology as SPEC CPU 2017.
For CloudSuite, we used the traces made available for the 2nd Cache
Replacement Competition (CRC-2)~\cite{CRC_2}. The traces include 4 4-core
applications with 6 distinct phases per application.

@@

\subsection{Prefetchers Simulated}
\label{Method-Prefetchers}
We compared PPF against three of the latest state of the art hardware-only
prefetchers: Best Offset Prefetcher (BOP), DRAM Aware - Access Map Pattern
Matching (DA-AMPM) \cite{DA_AMPM} and Signature Path Prefetcher (SPP). BOP was
the winner of 2nd Data Prefetching Championship. DA-AMPM is the enhanced
version of AMPM, modified to account for DRAM row buffer locality. SPP has
been shown to outperform BOP on SPEC CPU 2006 traces. For each of these, we
compare their speedups taking no prefetching as the baseline.

\subsection{Developing Features for PPF}
\label{Method-Features}
This section describes the intuition and analysis that went behind developing
the perceptron features. As noted earlier, we developed a set of 
nine features that allows the perceptron layer to correlate prefetching decision 
with the program behaviour. To study the correlation across each feature, 
we statistically examine the perceptron weights and try to 
interpret their distribution. \\ \\
%
\textbf{Global Pearson's Correlation}\newline 
Here we examine
the perceptron weights at the end of all trace execution by which time the
weights have settled to steady values. The weights obtained from running all
the SPEC CPU 2017 traces are concatenated. The feature with bulk of the
perceptron weights concentrated around 0 or small magnitude numbers show a
weak correlation with the prefetching outcome. On the other hand, features
with most of the weights saturated around highest value (+15) show a high
positive correlation and the features with weights close to the lowest value
(-16) show a strong negative correlation.

@@

We plot a histogram for each feature depicting weights distribution from -16
to +15 and generate the Pearson's correlation factor for that feature.
Pearson's factor is a numerical measure ranging from -1 to 1 of the degree of
linear correlation between two variables. The magnitude of Pearson's factor
gives the extent of correlation and the sign indicates whether it is a
positive correlation or a negative correlation. Values close to 0 suggest a
low correlation or at times, noisy data. A value of +1/-1 suggests a
perfectly linear positive / negative correlation respectively.

@@

Figure~\ref{Fig:Pvals} shows all the features which are finally used, 
arranged in the increasing order of their Pearson's factor. 
As can be seen 5 out of the 9 features provide a moderate to high 
correlation, with the magnitude of P-value > 0.6. The single most 
important feature, \textit{Confidence XOR Page address} helps provide
a correlation to prefetch outcome with a factor of 0.90. \\ \\
%
\textbf{Per Trace Correlation} \newline Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces. Here we give special attention to features
with low P-values in. Figure~\ref{Fig:Pvals_Trace} shows the
variation of P-values for three features : \textit{PC XOR Delta},
\textit{Signature XOR Delta} and 
\textit{PC XOR Depth}; across all the SPEC CPU 2017 traces. 
For simplicity, the traces are arranged in an
increasing order of contribution made by the feature. 
It can be seen that even features with a low
overall correlation provide useful correlation (magnitude > 0.5) for
a significant number of traces. This study motivated us to choose \textit{PC XOR DELTA}
over \textit{Last Signature} as it provided useful correlation in at least some of the
traces.\\ \\
%
\textbf{Trimming Features Using Cross-correlation} \newline Beside
providing interesting insights into prefetching behavior, P-value can
also be used for feature selection and prefetcher tuning.

@@

\footnotetext[1]{Components of Prefetch Table can be found in Table \ref{tab:PTable_metadata}}
\footnotetext[2]{The Reject Table does not need to maintain the useful bit as that only applies for prefetches that ultimately made through}

>>>>

\subsection{Performance Model}
\label{Method-Model}
We use the ChampSim~\cite{Champsim} simulator for the evaluation of
PPF against prior work techniques.  ChampSim is an enhanced version of
the framework that was used for the 2nd Data Prefetch Championship
(DPC-2)~\cite{DPC_2}, also used in the 2nd Cache Replacement
Competition (CRC2)~\cite{CRC_2}.  We model 1-core, 4-core, and 8-core
out-of-order machines.  The details of the configuration parameters
are summarized in Table~\ref{tab:Sim_params}.

@@

The block size is fixed at 64 bytes.  Prefetching is only triggered
upon L2 cache demand accesses but could be directed to the L2 or
last-level cache.  No L1 data level prefetching is done. The LRU
replacement policy is used on all levels of cache hierarchies. Branch
prediction is done using the perceptron branch
predictor~\cite{PerceptronPredictor}.  The page size is 4KB.  ChampSim
operates all the prefetchers strictly in the physical address space.

@@

%This is the first time that SPEC CPU 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
%We use the SPEC CPU 2017 benchmark suite~\cite{SPEC2017}. 
We use all the 20 workloads available in the SPEC CPU 2017
suite~\cite{SPEC2017}. Using the SimPoint~\cite{SimPoint} methodology,
we identified 95 different program segments of 1 Billion
instructions each.

\noindent \textbf{Single-core performance:} For single-core
simulations, we use the first 200 million instructions to warm-up the
microarchitectural structures and the next one billion instructions to
do detailed simulations and collect run-time statistics. We report the
IPC speedup over the baseline of no prefetching.  The final numbers
reported are the geometric mean of the weighted mean speedup achieved
per application using the SimPoint methodology.

\noindent \textbf{Multi-core performance:} For multi-application
workloads, we generate 100 random mixes and another 100 mixes from the
memory intensive subset of SPEC CPU 2017.  For 4-core workloads, 200
Million instructions are used for warm-up and additional 1 Billion
instruction simulated for collecting statistics.  Each CPU keeps
executing its workload till the last CPU completes one billion
instructions after warm-up.  For collecting IPC and other data, only
the first billion instructions are considered as the region of
interest.

@@

We repeat the same process for 8-core workloads, correspondingly with
16MB LLC. The only difference is that 20 million warm-up instructions
and 100 million full instructions are executed. This is done so as to
keep the simulation run-time within reasonable limits as a single
8-core mix takes up to 3 days to simulate one billion instructions.

\noindent \textbf{Validation:} We cross-validated our PPF model using
SPEC CPU 2006~\cite{SPEC2006} and CloudSuite~\cite{CloudSuite}
benchmarks. For single-core SPEC CPU 2006, we developed 94 simpoints
spread across all the 29 applications.  For multi-core, we followed
the same methodology as SPEC CPU 2017.  For CloudSuite, we used the
traces made available for the 2nd Cache Replacement Competition
(CRC-2)~\cite{CRC_2}. The traces include four 4-core applications with
six distinct phases per application.

@@

\subsection{Prefetchers Simulated}
\label{Method-Prefetchers}
We compared PPF against three of the latest, state of the art
hardware-only prefetchers: Best Offset Prefetcher (BOP), DRAM Aware -
Access Map Pattern Matching (DA-AMPM) \cite{DA_AMPM} and Signature
Path Prefetcher (SPP).  BOP was the winner of 2nd Data Prefetching
Championship. DA-AMPM is the enhanced version of AMPM, modified to
account for DRAM row buffer locality. SPP has been shown to outperform
BOP on SPEC CPU 2006 traces~\cite{SPP}. For each of these, we compare
their speedups taking the no prefetching case as the baseline.

\subsection{Developing Features for PPF}
\label{Method-Features}
This section describes the intuition and analysis that went behind
developing the perceptron features.  As noted earlier, we developed a
set of nine features that allow the perceptron layer to correlate
prefetching decision with the program behavior.  To study the
correlation across each feature, we statistically examine the
perceptron weights and try to interpret their distribution.

\noindent \textbf{Global Pearson's Correlation} Here we examine the
perceptron weights at the end of all trace execution by which time the
weights have settled to steady values. The weights obtained from
running all the SPEC CPU 2017 traces are concatenated.  Features with
a bulk of their perceptron weights concentrated around 0 or small
magnitude numbers show a weak correlation with the prefetching
outcome.  On the other hand, features with most of the weights
saturated around highest value (+15) show a high positive correlation
and the features with weights close to the lowest value (-16) show a
strong negative correlation.

@@

We plot a histogram for each feature depicting weights distribution
from -16 to +15 and generate the Pearson's correlation factor for that
feature.  Pearson's factor is a numerical measure ranging from -1 to 1
of the degree of linear correlation between two variables. The
magnitude of Pearson's factor gives the extent of correlation and the
sign indicates whether it is a positive correlation or a negative
correlation. Values close to 0 suggest a low correlation while a value
of +1/-1 suggests a perfectly linear positive / negative correlation
respectively.

@@

Figure~\ref{Fig:Pvals} shows all the features which are finally used,
arranged in the increasing order of their Pearson's factor.  As can be
seen 5 out of the 9 features provide a moderate to high correlation,
with the magnitude of P-value > 0.6. The single most important
feature, \textit{Confidence XOR Page address} helps provide
a correlation to prefetch outcome with a factor of 0.90.

\noindent \textbf{Per Trace Correlation} Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces. Here we give special attention to features
with low P-values in. Figure~\ref{Fig:Pvals_Trace} shows the variation
of P-values for three features : \textit{PC XOR Delta},
\textit{Signature XOR Delta} and \textit{PC XOR Depth}; across all the
SPEC CPU 2017 traces.  For simplicity, the traces are arranged in an
increasing order of contribution made by the feature.  It can be seen
that even features with a low overall correlation provide useful
correlation (magnitude > 0.5) for a significant number of traces. This
study motivated us to choose \textit{PC XOR DELTA} over \textit{Last
  Signature} as it provided useful correlation in at least some of the
traces.

\noindent \textbf{Trimming Features Using Cross-correlation} Beside
providing interesting insights into prefetching behavior, P-value can
also be used for feature selection and prefetcher tuning.

@@

\footnotetext[1]{Components of Prefetch Table can be found in Table
  \ref{tab:PTable_metadata}.}

\footnotetext[2]{The Reject Table does not need to maintain the useful
  bit as that only applies for prefetches that ultimately made
  through.}
